Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-08 write-ups/PLDI/paper.tex
Basically there with intro, given to Paul Thomson to read.

@@

\title{Responsive GPU Workgroup Management in the Presence of
  Persistent Kernels}

@@

@@

>>>>

\newcommand{\offerfork}{\mathsf{offer\_fork}}
\newcommand{\offerkill}{\mathsf{offer\_kill}}

@@

\title{Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs}

@@

\paragraph{The needs of irregular data-parallel algorithms}
Acceleration of general-purpose computations on graphics processing
units (GPUs) has tended to focus on \emph{regular} data-parallel
algorithms, for which work to be processed can be evenly split between
the workgroups that execute a GPU kernel ahead of time.  However, many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  There is growing interest in accelerating irregular
data-parallel algorithms on GPUs, particularly algorithms that
traverse and manipulate graphs, with several recent notable
successes~\cite{...}.

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups, to balance load and communicate intermediate
results.  For example, work stealing algorithms require each workgroup
to maintain a queue, which is typically protected by a mutex to enable
stealing by other workgroups.  As another example, breadth-first graph
traversal uses a frontier-based approach that requires a barrier
synchronization between all workgroups to achieve consensus that
exploration of a frontier has been accomplished.  \ADComment{Perhaps
  make the point that inter-workgroup synchronization is possible in
  very rare cases without blocking, e.g. threadfence reduction.}

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular
data-parallel algorithms accelerated on a GPU, this translates to
requiring fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if workgroup $A$ holds a mutex and workgroup $B$ is
spin-waiting for the mutex to be released, an unfair scheduler can
forever schedule $B$, so that $A$ never releases the mutex.
Similarly, an unfair scheduler can induce starvation if workgroup $A$
is at an inter-workgroup barrier awaiting the arrival of other
workgroups, by allowing A to spin forever so that the other workgroups
do not progress to the barrier.

\paragraph{Occupancy-bound execution}
Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
and HSA~\cite{...}---specify very few constraints on the manner in
which workgroups are scheduled.  In particular, they do \emph{not}
specify that workgroups must be fairly scheduled, and implementations
of these programming models do not provide fair scheduling in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.  If a kernel is executed
with more workgroups than there are compute units then at least two
workgroups must share a single compute unit.  The simplest way to
handle this is via an \emph{occupancy-bound} execution model, whereby
once a workgroup has commenced execution on a compute unit (is has
become \emph{occupant}), the workgroup has exclusive access to the
compute unit until it has finished execution, after which a subsequent
workgroup can be bound to the compute unit.  Experiments suggest that
the occupancy-bound execution model is widely implemented by today's
GPU devices and drivers~\cite{...}, and it is evidently a simple model
for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, because
they are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume that the target GPU device provides the occupancy-bound
execution model, which they exploit by scheduling no more workgroups
than there are available compute units~\cite{...}.  This works today
because \emph{today}'s GPUs seem to provide the occupancy-bound
execution model.

\paragraph{Resistance to occupancy-bound execution}

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable from the point-of-view of
GPU vendors and end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all workgroups, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.  Precisely because
of this problem, operating systems employ a GPU \emph{watchdog} to
heuristically detect and kill long-running computations on the GPU
being used for display rendering.  If this watchdog is disabled,
execution of a long-running kernel causes the display to freeze for
the duration of kernel execution.  Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets.  It is desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling requires being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts~\cite{CitePersonalCommunication}, is that
\textbf{(1)} GPU vendors will not commit to the occupancy-bound
execution model that they currently implement, due to the need for
multi-tasking and energy throttling, yet \textbf{(2)} GPU vendors will
not guarantee fair scheduling, due to the runtime overhead associated
with supporting full preemption of workgroups, e.g.\ based on
time-slicing.  Vendors instead wish to retain the essence of the
occupancy-bound execution model, which is simple to implement
efficiently, and provide preemption for the special cases associated
with multi-tasking for high-priority jobs (most crucially graphics)
and energy throttling.  In particular, they do not want to commit to
fair scheduling of a workgroup once it has been preempted.

\paragraph{Our proposal}
%
To summarise: blocking algorithms, such as most irregular algorithms,
demand a degree of fair scheduling of workgroups, but for good reasons
GPU vendors will not commit to the guarantees of even the
occupancy-bound execution model.

We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.  A GPU kernel
that requires workgroups to be fairly scheduled is identified as a
\emph{cooperative} kernel, and written using two additional language
primitives, $\offerkill$ and $\offerfork$, which must be placed
carefully by the programmer.

At a point where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$ to sacrifice itself to
the scheduler.  This indicates to the scheduler that the workgroup
would ideally continue with its execution, but that if required the
scheduler may destroy the workgroup for good: the cooperative kernel
is prepared to deal with either scenario.

At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates to the scheduler that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commencing execution at the $\offerfork$ program point, inheriting the
state of the workgroup that executed $\offerfork$.

The use of $\offerfork$ and $\offerkill$ creates a contract between
the scheduler and (the programmer of the) cooperative kernel, with
both functional and non-functional requirements.  Functionally: the
scheduler must guarantee that the workgroups executing a cooperative
kernel are fairly scheduled; the cooperative kernel must be robust to
workgroups leaving and joining the computation when the $\offerkill$
and $\offerfork$ functions, respectively, are called.
Non-functionally: a cooperative kernel must ensure that $\offerkill$
is executed frequently enough that the scheduler can accommodate
soft-real time constraints, e.g.\ allowing a smooth frame-rate for
graphics or allowing compute units to be powered down quickly when
required; the scheduler should respond to $\offerkill$ and
$\offerfork$ calls in a manner that avoids under-utilisation of
hardware resources by the cooperative kernel, e.g.\ the scheduler
should not kill workgroups unless they are required for another
purpose.

Cooperative kernels have several appealing properties.  First, they
accommodate the needs of blocking algorithms, including irregular
data-parallel algorithms: such algorithms can be implemented naturally
as cooperative kernels \ADComment{boast that we did a good set}, so
that fair scheduling between workgroups is guaranteed.  Second, they
have no impact on the development of regular (non-cooperative
kernels): these can be programmed exactly as they are now.  Third, the
programming model is backwards compatible: a non-blocking algorithm
implementation that works on today's GPU platforms can be upgraded to
use our language extensions for cooperative scheduling, and will
behave exactly as it does at present if the new $\offerkill$ and
$\offerfork$ functions are simply ignored.  Fourth, the programming
model is easy to implement on top of the occupancy-bound execution
model that current GPU platforms provide; we present a
proof-of-concept implementation on top of OpenCL 2.0 that requires no
special-purpose hardware or driver support.  Fifth, our results
suggest that cooperative kernels can provide efficient multi-tasking
of cooperative and non-cooperative tasks: our experiments across a
range of GPUs from two vendors show very good response times for
scheduling short, non-cooperative kernels concurrently with
long-running cooperative kernels.  These performance results are with
respect to our proof-of-concept, portable implementation; a vendor
specific implementation of cooperative kernels could likely yield
better performance still through device-specific driver and compiler
support.

In summary, our main contributions are:

\begin{enumerate}

\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms  \ADComment{Maybe big up to say that we give semantics to the new operations.}

\item A portable implementation of cooperative kernels on top of the OpenCL 2.0 programming model

\item The porting of \ADComment{TODO} irregular graph algorithms \ADComment{tighten} to the cooperative kernels model

\item \ADComment{The experiments}

\end{enumerate}

\ADComment{Text to maybe work in.}

In traditional co-operative multi-tasking, a task relinquishes control
and regains control in a stateful manner. We propose a lighter weight
model we believe to suitible for the type of blocking algorithms we
are aware of for GPUs.

The scheduler agrees to scheduler the workgroups fairly, but is at
liberty to eliminate workgroups for good when they offer themselves
to be killed. This allows the schduler to repond to changing depands
on the GPU, e.g. to implement energy throttling.

Prototype implementation of the programming model using OpenCL 2.0
advance features to build the scheudler and implement a clang based
compiler to transform input programs using the language extensions to
lower-level programs that interact with the scheduler. Our prototype
simulates multiple independent kernels into a mega-kernel. A vendor
specific implementation could achieve higher efficiency through
modications to the GPU driver that may take advantage of hardware
features. (e.g. mega kernel might not be good to compile). Its
reasonable to think our proof of concept provides a lower bound on how
well we can do.

We ported a large set of blocking algorithms to our programming
model. We evaluate co-operative scheduling with a cooperative kernel
and non-coperative kernel. GPUs from multiple vendors and access
various things.

Our results are promising.

In the evaluation, make a note that the applications we worked on
Namely if they use only a barrier or work-stealing idiom, are
straightforward to adopt to our programming model.

@@

\section{Old intro}
